Exoplanet population inference and the abundance \\ of Earth analogs from noisy, incomplete catalogs

### 2014-06-10 document/ms.tex
final version

The detection efficiency also depends on the geometric transit probability
$R_\star/a$.
Since we are modeling the distribution in the period--radius plane, we need to
compute the transit probability marginalized over stellar radius and mass.
This marginalized distribution scales only with the period of the orbit as
$\propto \period^{-2/3}$.
In theory, this marginalization should be over the \True\ distribution of
these parameters in the selected stellar catalog but we'll approximate it by
the empirical distribution; a reasonable approximation given the size of the
dataset.
At a period of 10 days\footnote{This period is chosen arbitrarily because the
power law only needs to be normalized at one point.}, the median transit
probability in the selected sample of stars is $5.061\%$ so we model the
transit probability\footnote{We are using the letter $Q$ to indicate
probabilities since we are already using $P$ to mean period.} as a function of
period as
\begin{eqnarray}
\transitprob (\period) &=&
    0.05061\,\left[\frac{\period}{10\,\mathrm{days}}\right]^{-2/3} \quad.
\end{eqnarray}
This expression is clearly only valid for $\period \gtrsim 1.4\,\mathrm{days}$
but the dataset that we are using (\citealt{petigura}) explicitly only
includes periods longer than five days so this is not a problem.
We're using the \emph{median} transit probability (instead of the mean)
because it is a more robust estimator in the presence of outliers but in our
experiments, the results do not seem to be very sensitive to this choice.

@@

Our extrapolation of the rate density to the location of Earth is more general
and conservative than any previously published method.
Using this method, we find a rate density of Earth analogs that is
inconsistent with the result published by \citet{petigura}.
This discrepancy can be largely attributed to the rigidity of the assumptions
about the period distribution.
Our extrapolation is less confident than previous measurements.
Again, this difference is due to the fact that we allow a much more flexible
extrapolation function.
\dfmtodo{This is another illustration that the use of flexible models makes
probabilistic data analysis more conservative.}

@@

\dfmtodo{%
Predicted number of observable Earth analogs in the \kepler\ dataset.
Low but not zero.
}

\dfmtodo{%
{\bf Advice:} \quad
Remind us that we want a representation of the full posterior (samples)
because that would let us do better things with the completeness (modeled as
r/R, \etc).
Must explicitly include priors too.
}

>>>>

The detection efficiency also depends on the geometric transit probability
$R_\star/a$.
Since we are modeling the distribution in the period--radius plane, we need to
compute the transit probability marginalized over stellar radius and mass.
This marginalized distribution scales only with the period of the orbit as
$\propto \period^{-2/3}$.
In theory, this marginalization should be over the \True\ distribution of
these parameters in the selected stellar catalog but we'll approximate it by
the empirical distribution; a reasonable approximation given the size of the
dataset.
At a period of 10 days\footnote{This period is chosen arbitrarily because the
power law only needs to be normalized at one point.}, the median transit
probability in the selected sample of stars is $5.061\%$ so we model the
transit probability\footnote{We are using the letter $Q$ to indicate
probabilities since we are already using $P$ to mean period.} as a function of
period as
\begin{eqnarray}\eqlabel{transitprob}
\transitprob (\period) &=&
    0.05061\,\left[\frac{\period}{10\,\mathrm{days}}\right]^{-2/3} \quad.
\end{eqnarray}
This expression is clearly only valid for $\period \gtrsim 1.4\,\mathrm{days}$
but the dataset that we are using (\citealt{petigura}) explicitly only
includes periods longer than five days so this is not a problem.
We're using the \emph{median} transit probability (instead of the mean)
because it is a more robust estimator in the presence of outliers but in our
experiments, the results do not seem to be very sensitive to this choice.

@@

Our extrapolation of the rate density to the location of Earth is more general
and conservative than any previously published method.
Using this method, we find a rate density of Earth analogs that is
inconsistent with the result published by \citet{petigura}.
This discrepancy can be largely attributed to the rigidity of the assumptions
about the period distribution.
Our extrapolation is less confident than previous measurements.
Again, this difference is due to the fact that we allow a much more flexible
extrapolation function.
This is another illustration that, against the standard data analysis
folklore, the correct use of flexible models is \emph{conservative}.

@@

Under the assumptions that we have made here, our inference of the occurrence
rate density of exoplanets places a probabilistic constraint on the number of
transiting Earth analogs in the existing \kepler\ dataset.
If we adopt the definition of ``Earth-like'' from \citet[][$200 \le
\period/\mathrm{day} < 400$ and $1 \le \radius/\radius_\oplus < 2$]{petigura},
and integrate the product inferred rate density function and the geometric
transit probability (\eqalt{transitprob}) over this bin, we find that there
should be
\begin{eqnarray}\eqlabel{ntransit}
N_{\oplus,\,\mathrm{transiting}} &=& 9.2_{-4.0}^{+5.9}
\end{eqnarray}
Earth-like exoplanets transiting the stars in the sample of Sun-like stars
chosen by \citet{petigura}.
This is an exciting result because it means that, if we can improve the
sensitivity of exoplanet search pipelines to small planets orbiting on long
periods then we should find some Earth analogs in the existing data.
Furthermore, because of the treatment of multiple transiting systems in the
catalog, the number in \eq{ntransit} is probably a lower limit on the expected
number of Earth-like exoplanets transiting Sun-like stars!

Some of the caveats on the results in this paper are due to assumptions made
for computational simplicity but a much more robust study would be possible
given a complete representation of the posterior probability function for the
physical parameters in the catalog.
The use of MCMC to fit models to observations is becoming standard practice in
astronomy and the results in many catalogs (including \citealt{petigura}) are
given as statistics computed on posterior samplings.
For the sake of hierarchical inferences like the method presented here, it
would be very useful if the authors of upcoming catalogs also published
samples from these distributions \emph{along with the value of their prior
function evaluated at each sample}.
In this spirit, we have released the results of this paper as posterior
samplings for the occurrence rate density function (\dfmtodo{ADD SOME URL}).
